### 2012-10-16 doc/prelim/document/chapters/introduction/introduction.tex
updates to parallel krylov performance section

\section{Hardware-Based Motivation}
\label{sec:hardware_motivation}
As leadership class machines move towards the exascale, new algorithms
must developed that leverage their strengths and adapt to their
shortcomings. Basic research is required now to advance methods in
time for these new machines to become operational. Organized work is
already moving forward in this area with the Department of Energy's
Advanced Scientific Computing Research office specifically allocating
funding for the next several years to research resilient solver
technologies for exascale facilities
\citep{u.s._department_of_energy_resilient_2012}. Based on the
language in this call for proposals, we can identify key issues for
which a set of robust, massively parallel Monte Carlo solvers could
provide a solution. As machines begin to operate at hundreds of
petaflops peak performance and beyond, restrictions on power
availability will require incredibly high levels of
concurrency. Furthermore, this drop in power consumption will mean
increased pressure on memory as memory per node is expected to
stagnate while cores per node is expected to increase. As the number
of cores increases, their clock speed is expected to stagnate or even
decrease to further reduce power consumption and manufacturing costs.

The end result of these hardware changes is that the larger numbers of
low-powered processors will be prone to both soft failures such as bit
errors in floating point operations and hard failures where the data
owned by that processor cannot be recovered. Because these failures
are predicted to be common, resilient solver technologies are required
overcome these events. With linear and nonlinear solvers based on
Monte Carlo techniques, such issues are alleviated by statistical
arguments. In the case of soft failures, isolated floating point
errors in Monte Carlo simulation are absorbed within tally statistics
while completely losing hardware during a hard failure is manifested
as a high variance event where some portion of the Monte Carlo
histories are lost. These stochastic methods are a paradigm shift from
current deterministic solver techniques that will suffer greatly from
the non-deterministic behavior expected from these exascale machines.

>>>>

\section{Hardware-Based Motivation}
\label{sec:hardware_motivation}
As leadership class machines move towards the exascale, new algorithms
must be developed that leverage their strengths and adapt to their
shortcomings. Basic research is required now to advance methods in
time for these new machines to become operational. Organized work is
already moving forward in this area with the Department of Energy's
Advanced Scientific Computing Research office specifically allocating
funding for the next several years to research resilient solver
technologies for exascale facilities
\citep{u.s._department_of_energy_resilient_2012}. Based on the
language in this call for proposals, we can identify key issues for
which a set of robust, massively parallel Monte Carlo solvers could
provide a solution. As machines begin to operate at hundreds of
petaflops peak performance and beyond, trends toward reduced energy
consumption will require incredibly high levels of concurrency to
achieve the desired computation rates. Furthermore, this drop in
power consumption will mean increased pressure on memory as memory per
node is expected to stagnate while cores per node is expected to
increase. As the number of cores increases, their clock speed is
expected to stagnate or even decrease to further reduce power
consumption and manufacturing costs.

The end result of these hardware changes is that the larger numbers of
low-powered processors will be prone to both soft failures such as bit
errors in floating point operations and hard failures where the data
owned by that processor cannot be recovered. Because these failures
are predicted to be common, resilient solver technologies are required
to overcome these events. With linear and nonlinear solvers based on
Monte Carlo techniques, such issues are alleviated by statistical
arguments. In the case of soft failures, isolated floating point
errors in Monte Carlo simulation are absorbed within tally statistics
while completely losing hardware during a hard failure is manifested
as a high variance event where some portion of the Monte Carlo
histories are lost. These stochastic methods are a paradigm shift from
current deterministic solver techniques that will suffer greatly from
the non-deterministic behavior expected from these exascale machines.

### 2012-10-16 doc/prelim/document/chapters/linear_problem/linear_problem.tex
working on adjoint section per spaniers work
updates to parallel krylov performance section

\chapter{Conventional Solution Methods for Linear Problems}
\label{ch:linear_problem}
The discretization of partial differential equations (\textit{PDEs})
through common methods such as finite differences
\citep{leveque_finite_2007}, finite volumes
\citep{leveque_finite_2002}, and finite elements
\citep{zienkiewicz_finite_2005} ultimately generates sets of coupled
equations in the form of matrix problems. In many cases, these
matrices are sparse, meaning that the vast majority of their
constituent elements are zero. This sparsity is due to the fact that
the influence of a particular grid element only expands as far as a
few of its nearest neighbors depending on the order of discretization
used and therefore coupling among variables in a particular discrete
equation in the system leads to a few non-zero entries. Because of the
natural occurrence of sparse matrices in common numerical methods many
iterative techniques have been developed to solve such systems. We
discuss here conventional stationary and projection methods for
solving sparse systems to provide the necessary background for the
remainder of this work. Details on the parallelization of conventional
methods are discussed.\footnote{The contents of this chapter,
  particularly those sections relating to projection methods and
  matrix analysis, are heavily based on Saad's text
  \citep{saad_iterative_2003}.}

@@

Given this minimization condition for the residual, we can form the
outline of an iterative projection method. Consider a matrix $\ve{V}$
to form a basis of $\mathcal{K}$ and a matrix $\ve{W}$ to form a basis
of $\mathcal{L}$. As $\boldsymbol{\delta} \in \mathcal{K}$ by
definition in Eq~(\ref{eq:linear_projection_step}), then
$\boldsymbol{\delta}$ can instead be rewritten as:
\begin{equation}
  \boldsymbol{\delta} = \ve{V}\ve{y},\ \forall \ve{y} \in \mathbb{R}^N:\,
  \label{eq:linear_delta_projection}
\end{equation}
where $\ve{V}$ \textit{projects} $\ve{y}$ onto $\mathcal{K}$. From the
orthogonality constraint in Eq~(\ref{eq:linear_projection_constraint})
it then follows that:
\begin{equation}
  \ve{y} = (\ve{W}^T\ve{A}\ve{V})^{-1}\ve{W}^T\ve{r}_0\:,
  \label{eq:linear_constraint_projection}
\end{equation}
where here the projection onto $\mathcal{K}$ is constrained by the
projection onto $\mathcal{L}$. Knowing this, we can then outline the
following iteration scheme for a projection method:
\begin{subequations}
  \begin{gather}
    \ve{r}^k = \ve{b} - \ve{A}\ve{x}^k\:,\\  
    \ve{y}^k = (\ve{W}^T\ve{A}\ve{V})^{-1}\ve{W}^T\ve{r}_0\:,\\
    \ve{x}^{k+1} = \ve{x}^k + \ve{V}\ve{y}^k\:,
  \end{gather}
  \label{eq:linear_projection_iteration}
\end{subequations}
where $\ve{V}$ and $\ve{W}$ are generated from the definitions of
$\mathcal{K}$ and $\mathcal{L}$ which are defined prior to each
iteration.

@@

communication is required. Next, process one gathers the components of
the global vector that it requires to complete its part of the vector
product. For this example, the components of matrix that will operate
on the global vector are zero, and therefore no vector elements are
required to be scattered from process 3 to process 1. Those matrix
elements that will act on the piece of the vector owned by process 2
are not all non-zero, and therefore we must gather the entire process
2 vector components onto process 1 to complete the multiplication.
Conversely, processes 2 and 3 must scatter its vector components that
are required by other processes (such as process 1) in order to
complete their pieces of the product. This then implies that these
domain connections for proper gather and scatter combinations must be
constructed a priori. These data structures are typically generated by
a data partitioning library. Mathematically, if we are performing a
global matrix-vector multiply of the form $\ve{A}\ve{x} = \ve{y}$,
then for this example on process 1 we have a sequence of local
matrix-vector multiplications: $\ve{A}_1\ve{x}_1 + \ve{A}_1\ve{x}_2 =
\ve{y}_1$. Here, some of the data is intrinsically local, and some
must be gathered from other processes using the partitioning data
structures.

\subsection{Parallel Performance Implications for Krylov Methods}
\label{subsec:projection_method_performance}
Knowing the parallel characteristics of the key operations we must
perform in order to implement Krylov methods, we can make a few
statements about parallel performance and implications for operation on
machines of increasing size.  For very large distributed machines,
global reduction operations required at several levels of Krylov
algorithms stand to reduce scalability and performance. Furthermore,
communication between adjacent domains in matrix-vector multiply
operations may also cause a bottleneck as the number of domains used
in simulation grows. The end result is that global data must be
collected and communicated. For scaling improvement, we seek a
reduction in these types of operations. In addition, these issues
become more prominent as the Krylov iterations progress, causing the
Krylov subspace to grow and the total number of operations needed to
orthogonalize that subspace to increase.

>>>>

\chapter{Conventional Solution Methods for Linear Systems}
\label{ch:linear_problem}
The discretization of partial differential equations (\textit{PDEs})
through common methods such as finite differences
\citep{leveque_finite_2007}, finite volumes
\citep{leveque_finite_2002}, and finite elements
\citep{zienkiewicz_finite_2005} ultimately generates sets of coupled
equations in the form of matrix problems. In many cases, these
matrices are sparse, meaning that the vast majority of their
constituent elements are zero. This sparsity is due to the fact that
the influence of a particular grid element only expands as far as a
few of its nearest neighbors depending on the order of discretization
used and therefore coupling among variables in a particular discrete
equation in the system leads to a few non-zero entries. Because of the
natural occurrence of sparse matrices in common numerical methods many
iterative techniques have been developed to solve such systems. We
discuss here conventional stationary and projection methods for
solving sparse systems to provide the necessary background for the
remainder of this work. Details on the parallelization of conventional
methods are discussed.\footnote{The contents of this chapter,
  particularly those sections relating to projection methods and
  matrix analysis, are heavily based on Saad's text
  \citep{saad_iterative_2003}.}

@@

Given this minimization condition for the residual, we can form the
outline of an iterative projection method. Consider a matrix $\ve{V}$
to form a basis of $\mathcal{K}$ and a matrix $\ve{W}$ to form a basis
of $\mathcal{L}$. As $\boldsymbol{\delta} \in \mathcal{K}$ by
definition in Eq~(\ref{eq:linear_projection_step}), then
$\boldsymbol{\delta}$ can instead be rewritten as:
\begin{equation}
  \boldsymbol{\delta} = \ve{V}\ve{y},\ \forall \ve{y} \in \mathbb{R}^N:\,
  \label{eq:linear_delta_projection}
\end{equation}
where $\ve{V}$ \textit{projects} $\ve{y}$ onto $\mathcal{K}$. From the
orthogonality constraint in Eq~(\ref{eq:linear_projection_constraint})
it then follows that:
\begin{equation}
  \ve{y} = (\ve{W}^T\ve{A}\ve{V})^{-1}\ve{W}^T\ve{r}_0\:,
  \label{eq:linear_constraint_projection}
\end{equation}
where here the projection onto $\mathcal{K}$ is constrained by the
projection onto $\mathcal{L}$. Knowing this, we can then outline the
following iteration scheme for a projection method:
\begin{subequations}
  \begin{gather}
    \ve{r}^k = \ve{b} - \ve{A}\ve{x}^k\:,\\  
    \ve{y}^k = (\ve{W}^T\ve{A}\ve{V})^{-1}\ve{W}^T\ve{r}^k\:,\\
    \ve{x}^{k+1} = \ve{x}^k + \ve{V}\ve{y}^k\:,
  \end{gather}
  \label{eq:linear_projection_iteration}
\end{subequations}
where $\ve{V}$ and $\ve{W}$ are generated from the definitions of
$\mathcal{K}$ and $\mathcal{L}$ and are updated prior to each
iteration.

@@

communication is required. Next, process 1 gathers the components of
the global vector owned by the other two processes that it requires to
complete its part of the vector product. For this example, the
components of matrix owned by process 1 that will operate on the
global vector components owned by process 3 are zero, and therefore no
vector elements are required to be scattered from process 3 to process
1. Those matrix elements owned by process 1 that will act on the piece
of the vector owned by process 2 are not all non-zero, and therefore
we must gather the entire process 2 vector components onto process 1
to complete the multiplication.  Conversely, processes 2 and 3 must
scatter their vector components that are required by other processes
(such as process 1) in order to complete their pieces of the
product. This then implies that these domain connections for proper
gather and scatter combinations must be constructed a priori. These
data structures are typically generated by a data partitioning
library. Mathematically, if we are performing a global matrix-vector
multiply of the form $\ve{A}\ve{x} = \ve{y}$, then for this example on
process 1 we have a sequence of local matrix-vector multiplications:
$\ve{A}_1\ve{x}_1 + \ve{A}_1\ve{x}_2 = \ve{y}_1$ as annotated by the
notation in Figure~\ref{fig:matvec_proc_1}. Here, some of the data is
intrinsically local, and some must be gathered from other processes
using the partitioning data structures.

\subsection{Parallel Performance Implications for Krylov Methods}
\label{subsec:projection_method_performance}
Knowing the parallel characteristics of the key operations we must
perform in order to implement Krylov methods, we can make a few
statements about parallel performance and implications for operation
on machines of increasing size. Reconsider the matrix and vector
operations required to implement Algorithm~\ref{alg:gmres}. For very
large distributed machines, the global reduction operations required
at several levels of Krylov algorithms stand to reduce scalability and
performance. In the case of GMRES, these reductions include vector
norm operations and the inner products required for basis
orthogonalization. Furthermore, communication between adjacent domains
in matrix-vector multiply operations may also cause a bottleneck as
the number of domains used in a simulation grows and the number of
processors participating in the gather/scatter sequence requires a
large communication bandwidth. The end result is that global data must
be collected and communicated. For scaling improvement, we seek a
reduction in these types of operations. In addition, these issues
become more prominent as the Krylov iterations progress, causing the
Krylov subspace to grow and the total number of operations needed to
orthogonalize that subspace to increase.

As an example of these performance implications in practice, in a 2001
work, Gropp and colleagues presented results on fluid dynamic
simulations that heavily leveraged Krylov methods in their solution
schemes \citep{gropp_high-performance_2001}. In this follow-on to
their 1997 Bell Prize-winning work, part of their analysis included
identifying parallel scalability bottlenecks generated by solver
implementations in a strong scaling exercise where the number of
processors was increased with respect to a fixed global problem
size. Gropp's observations show that for their particular hardware, a
distributed memory machine similar to modern architectures, that
global reduction operations did not impede scalability, meaning that
the global reduction operation occupied approximately the same
percentage of compute time independent of the number of processors
used. Rather, it was the gather/scatter operations required to
communicate data to neighboring processors that reduced performance
with an increasing percentage of compute time consumed by these
operations as processor count was increased. Furthermore, it was noted
that this reduction in scaling was a product of poor algorithmic
strong scaling rather than hardware or implementation related issues
as the algorithm requires more data to be scattered/gathered as the
number of processors and therefore computational domains increased. In
the case of a weak scaling exercise, we would instead expect this
percentage to exhibit a more desirable behavaior of remaining constant
for gather/scatter operations as problem size would be scaled with the
number of processors.
